### 2022-06-19 chapter6/chapter.tex
Ch7 2nd draft

%!TEX root = ../thesis.tex
\chapter{Numerical investigations\label{sec: numerics}}
We have now established multiple schemes for approximating fluid queues which are suitable for approximating the performance measures of fluid-fluid queues derived in \cite{bo2014}. In this chapter we numerically investigate various aspects of the approximation schemes. Throughout, we compare the QBD-RAP scheme (from Chapter~\ref{sec: construction and modelling}), the discontinuous Galerkin scheme (as described in Chapter~\ref{ch:galerkin}), and the spatially-coherent uniformisation scheme introduced by \cite{bo2013}. %Recall that the uniformisation scheme is equivalent to a discontinuous Galerkin scheme with a constant basis function in each cell, which is also equivalent to the finite-volume scheme with an upwind flux. 
Recall that, due to their stochastic interpretation, the uniformisation and QBD-RAP schemes are positivity preserving.

The main focus of our numerical experiments is on the convergence properties as the number of basis functions is increased. For simplicity, the number of basis functions is kept constant across all cells. We refer to the number of basis functions on a cell \(k\) as the \emph{dimension} and consider dimensions \(p=2r+1\) for \(r=0,1,2,3,...,10\). For discontinuous Galerkin scheme, increasing the dimension of the scheme means increasing the number of polynomial basis functions used to approximate the solution within each cell. E.g.~if we use 3 basis functions in the discontinuous Galerkin scheme, we approximate the solution by a quadratic on each cell. For the QBD-RAP scheme, increasing the dimension of the scheme means increasing the order of the CME distribution used to construct the scheme. To make a comparable equivalent for the uniformisation scheme with a fixed cell size we divide each cell into smaller sub-cells over which we approximate the solution. I.e.~for a dimension \(p\) uniformisation scheme we divide each cell into \(p\) sub-cells. Equivalently, we may think of a dimension \(p\) uniformisation scheme as using \(p\) piecewise constant functions to approximate the solution on each cell. For all schemes (DG, QBD-RAP and uniformisation), if we construct an order \(p\) approximation, there are \(K\) cells, \(N\) phases and \(c_i\neq 0\) for all \(i\in\calS\), then the resulting approximation to the generator \(\mathbb B\) is a matrix of dimension \(pKN + N\). In this sense each approximation scheme leads to matrices of the same size (although not necessarily the same number of non-zero elements).

To keep the content of this chapter contained, we do not investigate all aspects of the schemes. For each numerical experiment we keep the cell size, \(\Delta\), fixed for the DG and QBD-RAP schemes. As part of deriving a discontinuous Galerkin scheme, one needs to choose the \emph{numerical flux} which is used to approximate the transition of density from one cell to the next \citep{nodalDGBook}. We investigate schemes with an upwind flux only. For schemes which require us to integrate over time we do not investigate the stability of the schemes with respect to the \(t\)-step-size, or the time-integration scheme itself (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value to obey a certain stability criterion (a CFL-like condition, \citep[Section~4.8]{nodalDGBook}). Moreover, we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages (\cite{sr2002}, \cite{nodalDGBook}~Section~5.7, see also Section~\ref{sec: time integration}), which claims to introduce no more oscillations into the solution as we integrate over time.

For some numerical experiments we implement the discontinuous Galerkin scheme with and without a slope-limiter implemented (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}). The slope limiter avoids oscillations and negative solutions. However, we note that in the context of approximating the operator \(\mathbb\Psi\) for a fluid-fluid queue, it is not obvious how one might apply the concept of slope-limiting, other than to post-process the solution with a limiter or filter. Even when slope limiting is possible, the resultant approximation is at best linear around discontinuities. Moreover, there is a computational cost in applying a limiter, which could be significant, particularly if many initial conditions are to be considered. In contrast, for the uniformisation and QBD-RAP the approximate solution is guaranteed to produce positive probabilities without any post-processing or extra computation.

Where a slope limiter is implemented, we implement the \emph{Generalised MUSCL} limiter (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}). We investigate two schemes with slope limiting. In the first scheme, which we refer to the \emph{DG-lim} scheme, we use a DG scheme of dimension \(p\) and apply the Generalised MUSCL slope limiter as described in Section~\ref{sec: slope limiting}. In the second scheme, which we refer to as the \emph{DG-lin-lim} scheme, we follow the construction described in~\ref{sec: anoth pos pres}, which is effectively a DG scheme with cells of width \(2\Delta/(p+1)\), for \(p=1,3,5,...,21\), and a basis of dimension \(2\) (linear) on each cell, with an upwind flux, and we apply the Generalised MUSCL limiter. Thus, the DG-lin-lim scheme uses \(p+1\), \(p=1,3,5,...,21\), basis functions to represent the solution on each of the original cells \(\calD_k\), and we define the dimension of the scheme to be \(p+1\). We do not investigate filtering for the DG scheme.

The performance of the discontinuous Galerkin scheme has been well-studied in some contexts (\cite{c99}, \cite{nodalDGBook}~Section~5.5), and it is well-known that the discontinuous Galerkin scheme performs remarkably well on problems with smooth solutions. Here, we mostly focus on investigating the numerical performance of the schemes on problems with non-smooth solutions. The purpose of this is to emphasise the positivity-preserving properties of the QBD-RAP scheme. In the stochastic modelling community it is very common to have problems with discontinuities, such as a non-smooth initial conditions. As we will see (in Section~\ref{sec: transient approx}), even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities or non-smooth behaviour into the problem. A very specific set of conditions on the initial density and point masses must hold for the transient distribution to remain continuous as it evolves over time \citep{bo2014,bop2020}. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues (see, for example, Section~\ref{sec: return approx}). Limiting distributions of fluid queues are smooth, however.

As we shall see, if the solution is smooth, then the DG scheme is highly-effective and exhibits rapid convergence to the solution as the number of polynomial basis functions is increased. However, when the solution is non-smooth oscillations and negative approximations may occur when using the discontinuous Galerkin scheme, leading to nonsense solutions. In these cases, the QBD-RAP and DG-lin-lim approximations perform relatively well compared to the other positivity preserving schemes investigated.

Lastly, before we start the numerical investigation, note that all the following error plots are on a \(\log_{10}\)-\(\log_{10}\) scale. If the error function takes the form \(error = \beta_0^* p^{\beta_1}\), where \(p\) is the dimension of the scheme, then the \(\log_{10}\)-\(\log_{10}\) will show a linear trend; \(\log_{10}(error) = \beta_0 + \beta_1 \log_{10}(p)\), where \(\beta_0=\log_{10}(\beta_0^*)\). Where relevant, to estimate the asymptotic rate of convergence, \(\beta_1\), we estimate the of the line \(\log_{10}(error) = \beta_0 + \beta_1 \log_{10}(p)\) using ordinary least squares and the last eight data points on each plot. We plot the estimated lines of best fit and quote their slope to the right of the plots.

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance. This allows us to instrument the performance of the reconstruction methods without considering a specific model or any dynamics of the problem. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which allows us to instrument the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates the ability of the schemes to approximate the stationary distribution of the model, Section~\ref{sec: transient approx} investigates the ability of the schemes to approximate the transient distribution of the model for two initial conditions, and Section~\ref{sec: return approx} investigates the ability of the schemes to approximate the first hitting time of the fluid level on the boundary of the interval \([0,1]\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num} and approximate the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{Y(t)\}\) first returns to \(0\).

@@

% 3. STATIONARY DISTRIBUTIONS
	% i. SIMPLER
	% ii. MORE COMPLEX

@@

\section{Function approximation/reconstruction}\label{sec: recon num}
We start our investigation by looking at how well the schemes perform at approximating an initial condition. By approximating the initial condition only, we aim to instrument the performance of the approximation schemes without any dynamics.

For the discontinuous Galerkin method, we project the initial condition on to the set of polynomial functions which define the scheme, for the spatially-coherent uniformisation scheme, we look at the sub-cell averages of the initial condition, and for the QBD-RAP scheme, we compute the initial vector for the approximation, then reconstructing the solutions as described in Sections~\ref{sec: initial conditions} and~\ref{sec: closing}. For the purposes of approximation and reconstruction for the QBD-RAP scheme we must orientate the initial condition in a positive or negative direction; here, we suppose that the initial conditions belong to a positive phase. First we investigate three closing vectors we can use for the reconstruction in the QBD-RAP scheme. From the investigation, we decide to use the closing vector (\ref{eqn:5462}) throughout the rest of the chapter.

For this investigation we consider approximating initial conditions over a single cell of width \(\Delta = 1\). To numerically evaluate integrals arising in the approximation step (inner products and cell averages) we use a trapezoidal rule with 10,001 function evaluations. Similarly, we use 10,001 function evaluations to approximate \(L^p\) errors, and also as a finite set of points over which to compute the KS statistics.

@@

\subsection{QBD-RAP closing operators}\label{sec: closing vecs numerics}
The three closing operators we investigate are the following. The unnormalised closing vector in Equation~(\ref{eqn: alal}). A normalised version of the closing operator in Equation~(\ref{eqn:546}), with the normalised version given by (\ref{eqn: nonlin}). We refer to the closing vector in (\ref{eqn: nonlin}) as the \emph{naive normalised} closing vector. Technically we have not yet proved that the \emph{naive normalised} closing vector leads to a convergent approximation scheme. The third closing vector we investigate is that in Equation~(\ref{eqn:5462}) and we refer to this as the \emph{normalised closing operator}.

\begin{example}\label{ex: another example}
	First consider approximating the initial condition with density function \(2\times 1(x<0.5)\). In the left-hand side of Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} we plot the Kolmogorov-Smirnov error between the true CDF and the CDF constructed via the QBD-RAP approximation with the three different closing vectors. Interestingly, in this case, the unnormalised closing operator outperforms the other two at low dimensions. In this case, it just so happens that the error due to truncation of the unnormalised reconstruction constructively interferes with other errors to cause the error to be lower at low dimensions for the unnormalised scheme.

	Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also shows that the naive normalised and normalised reconstructions perform very similarly -- this is the case throughout much of this subsection. The difference between the naive normalised and normalised reconstructions is how they treat mass in the tail of a matrix exponential distribution (from \(2\Delta\) onward). Intuitively, there should be very little mass in the tail of the distribution as we are using concentrated matrix exponential distributions in the reconstruction.

>>>>

%!TEX root = ../thesis.tex
\chapter{Numerical investigations\label{sec: numerics}}
We have now established multiple schemes for approximating fluid queues which are suitable for approximating the performance measures of fluid-fluid queues derived in \cite{bo2014}. This chapter investigates, numerically, various aspects of the approximation schemes. Throughout, we compare the QBD-RAP scheme (from Chapter~\ref{sec: construction and modelling}), the discontinuous Galerkin scheme (as described in Chapter~\ref{ch:galerkin}, with and without slope limiting), and the \emph{spatially-coherent uniformisation} scheme introduced by \cite{bo2013}.

The performance of the discontinuous Galerkin scheme has been well-studied in some contexts (\cite{c99}, \cite{nodalDGBook}~Section~5.5), and it is well-known that the discontinuous Galerkin scheme performs remarkably well on problems with smooth solutions. Here, we mostly focus on investigating the numerical performance of the schemes on problems with non-smooth solutions. The purpose is to emphasise the probabilistic interpretation and positivity-preserving properties of the QBD-RAP scheme. In the stochastic modelling community it is common to have problems with discontinuities, such as non-smooth initial conditions, hence the type of problems we investigate in this chapter are of relevance. As we will see (in Section~\ref{sec: transient approx}), even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities or non-smooth behaviour into the problem. A very specific set of conditions on the initial density and point masses must hold for the transient distribution to remain continuous as it evolves over time \citep{bo2014,bop2020}. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues (see, for example, Section~\ref{sec: ffq num}). Limiting distributions of fluid queues are smooth, however.

The experiments demonstrate that the DG scheme can perform extremely well when the true solution is sufficiently smooth. However, when discontinuities are present in the true solution, the DG scheme can lead to approximate solutions which are oscillatory and this may lead to approximate solutions which violate the axioms of probability, such as approximations to PDFs which take negative values (equivalently, approximations to CDFs which are non-increasing), and in the worst cases, approximations to CDFs which take values outside \([0,1]\). Hence, there is a need for schemes which ensure the axioms of probability are maintained.

By design, the QBD-RAP scheme, uniformisation scheme, and DG schemes with slope limiting, produce approximations which obey the axioms of probability, i.e. produce approximations to PDFs which are non-negative, and CDFs which are non-decreasing, and \([0,1]\)-valued.

We implement two slope limited DG schemes, the DG-lim and DG-lin-lim scheme, which we describe in more detail in the next section. The numerical experiments demonstrate that the DG-lim scheme does not lead to a viable approximation scheme in the presence of discontinuities, while the DG-lin-lim scheme results in a relatively performant approximation scheme for the discontinuous problems considered here. In general, the QBD-RAP scheme performs similarly to the DG-lin-lim scheme, while the uniformisation scheme is a viable approximation method, but produces larger errors than the QBD-RAP and DG-lin-lim scheme, for the examples considered here.

Furthermore, when slope limiting is used, the resultant approximation is at best linear around discontinuities. Moreover, there is a computational cost in applying a limiter (or filter), which could be significant, particularly if many initial conditions are to be considered. In contrast, for the uniformisation and QBD-RAP, the approximate solution is guaranteed to produce positive probabilities without any post-processing or extra computation.

Lastly, for the DG scheme I know of no rigorous argument which shows that we can use the DG approximation to the generator of a fluid queue to approximate performance measures of fluid-fluid queues, such as the operator \(\mathbb \Psi\). Here, we rely on the intuition that substituting the DG approximation for the true operator \emph{should} give reasonable results, and further, \emph{assume} that the resulting approximations are polynomial. For example this means that, other than being the solution to a matrix Riccati equation, the DG approximations to \(\mathbb \Psi\) that we compute in this chapter are not actually rigorously defined.

On the other hand, using the stochastic approximations (QBD-RAP of uniformisation), results in rigorously defined operators as proved in \cite{bgnp2021}.

In summary, the QBD-RAP and uniformisation schemes are the only two schemes which result in rigorously defined approximations, preserve the axioms of probability, and, of the two, the numerical experiments demonstrate that the QBD-RAP methods is superior.

\section{Preliminaries}
Before presenting the results of the experiments, we need to describe some aspects of the approximation schemes in more detail, which we now proceed to do.

The main focus of the numerical experiments is on the error as the number of basis functions per cell is increase. In the numerical experiments the number of basis functions used to approximate the solutions is the same on all cells and the number of basis functions on a cell (and hence all cells) is referred to as the \emph{dimension} of the scheme. For the discontinuous Galerkin scheme, increasing the dimension of the scheme means increasing the number of polynomial basis functions used to approximate the solution within each cell. For example, if we use 3 basis functions in the discontinuous Galerkin scheme, we approximate the solution by 3 linearly independent quadratics on each cell and hence the approximation is a quadratic on each cell. For the QBD-RAP scheme, increasing the dimension of the scheme means increasing the order of the CME distribution used to construct the scheme. To make a comparable equivalent for the uniformisation scheme, we divide each cell into smaller sub-cells over which we approximate the solution. That is, for a dimension \(p\) uniformisation scheme we divide each cell into \(p\) sub-cells. Equivalently, we may think of a dimension \(p\) uniformisation scheme as using \(p\) piecewise constant functions to approximate the solution on each cell. For all schemes (DG, QBD-RAP and uniformisation), if we construct a dimension \(p\) approximation, there are \(K\) cells, \(N\) phases and \(c_i\neq 0\) for all \(i\in\calS\), then the resulting approximation to the generator \(\mathbb B\) is a square matrix of dimension \(pKN + N\).\footnote{There are \(pK\) basis functions for each phase which approximate the solution on the interior of the domain, hence \(pKN\) basis functions in total, plus an additional \(N\) dimensions to capture boundary conditions.} In this sense each approximation scheme leads to matrices of the same order (although not necessarily the same number of non-zero elements).

We implement the discontinuous Galerkin scheme with and without a slope-limiter (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}). The slope limiter stops oscillations appearing in the approximations and hence also avoids non-probabilistic approximations to PDFs and CDFs. We investigate two slope-limited schemes. In the scheme we refer to the \emph{DG-lim} scheme, we use a DG scheme of dimension \(p\) and apply the Generalised MUSCL slope limiter as described in Section~\ref{sec: slope limiting}. In the \emph{DG-lin-lim} scheme, we follow the construction described in Section~\ref{sec: anoth pos pres}, which is effectively a DG scheme with cells of width \(2\Delta/(p+1)\), for \(p=1,3,5,...,21\), and a basis of dimension \(2\) (linear) on each cell, with an upwind flux, and we apply the Generalised MUSCL limiter. Thus, the DG-lin-lim scheme uses \(p+1\), \(p=1,3,5,...,21\), basis functions to represent the solution on each of the original cells \(\calD_k\), and we define the dimension of the scheme to be \(p+1\). In the presence of discontinuities, slope-limiting forces the approximation to be linear, i.e.~an order~1 polynomial, around the discontinuity. For DG schemes of dimension greater than 2 with a slope limiter and in the presence of discontinuities, the ability of the scheme to represent higher-order polynomials is not well-utilised as the slope limiter forces the solution to be linear anyway. The DG-lin-lim scheme recognises this and instead of using higher-order polynomials, uses piecewise-linear functions and smaller cell widths to approximate the solution.

The concept of slope-limiting originates in the context of numerically integrating (typically with respect to a \emph{time} variable) a so-called \emph{semi-discrete system}, as we do in Section~\ref{sec: transient approx}. However, approximating the operator \(\mathbb\Psi\) of a fluid-fluid queue amounts to solving an implicit matrix-equation and there is no concept of numerical integration with respect to time. Hence, the concept of slope-limiting, it is not directly transferable. One could process the initial condition and post-process the solution with a limiter. However, there are possibly better choices for this post-processing, such as filtering, \citep[Section~5.6.1]{nodalDGBook}. One could also view the composition of the DG operator and the limiting operator as a one, but the result is a non-linear operator which is of no use for simplifying the Riccati Equation which characterises \(\mathbb \Psi\). Considering there is no theory on the use of the DG scheme to approximate \(\mathbb \Psi\) and hence the resultant approximation is not rigorously-defined, it is not known if the use of a limiter in this context is possible.

To keep the content of this chapter contained, we do not investigate all aspects of the schemes. For each numerical experiment we keep the number of cells, \(K\) (equivalently, the cell size, \(\Delta\)), fixed for the DG and QBD-RAP schemes. As part of deriving a discontinuous Galerkin scheme, one needs to choose the \emph{numerical flux} which is used to approximate the transition of density from one cell to the next \citep{nodalDGBook}. We investigate schemes with an upwind flux only. We do not investigate filtering for the DG scheme.

For schemes which require us to integrate over time we do not investigate the stability of the schemes with respect to the \(t\)-step-size, or the time-integration scheme itself (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value to obey a certain stability criterion (a CFL-like condition, \citep[Section~4.8]{nodalDGBook}). Moreover, we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages (\cite{sr2002}, \cite{nodalDGBook}~Section~5.7, see also Section~\ref{sec: time integration}), which claims to introduce no more oscillations into the solution as we integrate over time.

Where a slope limiter is implemented, we implement the \emph{Generalised MUSCL} limiter (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}).

In this chapter, all the error plots are on a \(\log_{10}\)-\(\log_{10}\) scale. If the error function takes the form \(error = \beta_0^* p^{\beta_1}\), where \(p\) is the dimension of the scheme, then the \(\log_{10}\)-\(\log_{10}\) will show a linear trend since \(\log_{10}(error) = \beta_0 + \beta_1 \log_{10}(p)\), where \(\beta_0=\log_{10}(\beta_0^*)\). Where relevant, to estimate the asymptotic rate of convergence, \(\beta_1\), we estimate the slope of the line \(\log_{10}(error) = \beta_0 + \beta_1 \log_{10}(p)\) using ordinary least squares and the last eight data points of each line. The estimated linear trends are plotted and the slope of the line is displayed to the right of the plot.

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance which allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which means that the solution is known, and we do not need to resort to simulation as a \emph{ground-truth}. Furthermore, the travelling wave problems allow us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates approximations to the limiting distribution of the queue, Section~\ref{sec: transient approx} investigates approximations to the transient distribution of the queue for two initial conditions, and Section~\ref{sec: hit approx} investigates approximations the first hitting time of the fluid level on the set \(\{0,1\}\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num}, compute approximations to \(\mathbb \Psi\) and ultimately, approximate the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{Y(t)\}\) first returns to \(0\).

% \chapter{Old}
% We have now established multiple schemes for approximating fluid queues which are suitable for approximating the performance measures of fluid-fluid queues derived in \cite{bo2014}. In this chapter we numerically investigate various aspects of the approximation schemes. Throughout, we compare the QBD-RAP scheme (from Chapter~\ref{sec: construction and modelling}), the discontinuous Galerkin scheme (as described in Chapter~\ref{ch:galerkin}), and the spatially-coherent uniformisation scheme introduced by \cite{bo2013}. %Recall that the uniformisation scheme is equivalent to a discontinuous Galerkin scheme with a constant basis function in each cell, which is also equivalent to the finite-volume scheme with an upwind flux. 
% Recall that, due to their stochastic interpretation, the uniformisation and QBD-RAP schemes are positivity preserving.

% The main focus of our numerical experiments is on the convergence properties as the number of basis functions is increased. For simplicity, the number of basis functions is kept constant across all cells. We refer to the number of basis functions on a cell \(k\) as the \emph{dimension} and consider dimensions \(p=2r+1\) for \(r=0,1,2,3,...,10\). For discontinuous Galerkin scheme, increasing the dimension of the scheme means increasing the number of polynomial basis functions used to approximate the solution within each cell. E.g.~if we use 3 basis functions in the discontinuous Galerkin scheme, we approximate the solution by a quadratic on each cell. For the QBD-RAP scheme, increasing the dimension of the scheme means increasing the order of the CME distribution used to construct the scheme. To make a comparable equivalent for the uniformisation scheme with a fixed cell size we divide each cell into smaller sub-cells over which we approximate the solution. I.e.~for a dimension \(p\) uniformisation scheme we divide each cell into \(p\) sub-cells. Equivalently, we may think of a dimension \(p\) uniformisation scheme as using \(p\) piecewise constant functions to approximate the solution on each cell. For all schemes (DG, QBD-RAP and uniformisation), if we construct an order \(p\) approximation, there are \(K\) cells, \(N\) phases and \(c_i\neq 0\) for all \(i\in\calS\), then the resulting approximation to the generator \(\mathbb B\) is a matrix of dimension \(pKN + N\). In this sense each approximation scheme leads to matrices of the same size (although not necessarily the same number of non-zero elements).

% To keep the content of this chapter contained, we do not investigate all aspects of the schemes. For each numerical experiment we keep the cell size, \(\Delta\), fixed for the DG and QBD-RAP schemes. As part of deriving a discontinuous Galerkin scheme, one needs to choose the \emph{numerical flux} which is used to approximate the transition of density from one cell to the next \citep{nodalDGBook}. We investigate schemes with an upwind flux only. For schemes which require us to integrate over time we do not investigate the stability of the schemes with respect to the \(t\)-step-size, or the time-integration scheme itself (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value to obey a certain stability criterion (a CFL-like condition, \citep[Section~4.8]{nodalDGBook}). Moreover, we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages (\cite{sr2002}, \cite{nodalDGBook}~Section~5.7, see also Section~\ref{sec: time integration}), which claims to introduce no more oscillations into the solution as we integrate over time.

% For some numerical experiments we implement the discontinuous Galerkin scheme with and without a slope-limiter implemented (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}). The slope limiter avoids oscillations and negative solutions. However, we note that in the context of approximating the operator \(\mathbb\Psi\) for a fluid-fluid queue, it is not obvious how one might apply the concept of slope-limiting, other than to post-process the solution with a limiter or filter. Even when slope limiting is possible, the resultant approximation is at best linear around discontinuities. Moreover, there is a computational cost in applying a limiter, which could be significant, particularly if many initial conditions are to be considered. In contrast, for the uniformisation and QBD-RAP the approximate solution is guaranteed to produce positive probabilities without any post-processing or extra computation.

% Where a slope limiter is implemented, we implement the \emph{Generalised MUSCL} limiter (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}). We investigate two schemes with slope limiting. In the first scheme, which we refer to the \emph{DG-lim} scheme, we use a DG scheme of dimension \(p\) and apply the Generalised MUSCL slope limiter as described in Section~\ref{sec: slope limiting}. In the second scheme, which we refer to as the \emph{DG-lin-lim} scheme, we follow the construction described in~\ref{sec: anoth pos pres}, which is effectively a DG scheme with cells of width \(2\Delta/(p+1)\), for \(p=1,3,5,...,21\), and a basis of dimension \(2\) (linear) on each cell, with an upwind flux, and we apply the Generalised MUSCL limiter. Thus, the DG-lin-lim scheme uses \(p+1\), \(p=1,3,5,...,21\), basis functions to represent the solution on each of the original cells \(\calD_k\), and we define the dimension of the scheme to be \(p+1\). We do not investigate filtering for the DG scheme.

% The performance of the discontinuous Galerkin scheme has been well-studied in some contexts (\cite{c99}, \cite{nodalDGBook}~Section~5.5), and it is well-known that the discontinuous Galerkin scheme performs remarkably well on problems with smooth solutions. Here, we mostly focus on investigating the numerical performance of the schemes on problems with non-smooth solutions. The purpose of this is to emphasise the positivity-preserving properties of the QBD-RAP scheme. In the stochastic modelling community it is very common to have problems with discontinuities, such as a non-smooth initial conditions. As we will see (in Section~\ref{sec: transient approx}), even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities or non-smooth behaviour into the problem. A very specific set of conditions on the initial density and point masses must hold for the transient distribution to remain continuous as it evolves over time \citep{bo2014,bop2020}. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues (see, for example, Section~\ref{sec: return approx}). Limiting distributions of fluid queues are smooth, however.

% As we shall see, if the solution is smooth, then the DG scheme is highly-effective and exhibits rapid convergence to the solution as the number of polynomial basis functions is increased. However, when the solution is non-smooth oscillations and negative approximations may occur when using the discontinuous Galerkin scheme, leading to nonsense solutions. In these cases, the QBD-RAP and DG-lin-lim approximations perform relatively well compared to the other positivity preserving schemes investigated.

% Lastly, before we start the numerical investigation, note that all the following error plots are on a \(\log_{10}\)-\(\log_{10}\) scale. If the error function takes the form \(error = \beta_0^* p^{\beta_1}\), where \(p\) is the dimension of the scheme, then the \(\log_{10}\)-\(\log_{10}\) will show a linear trend; \(\log_{10}(error) = \beta_0 + \beta_1 \log_{10}(p)\), where \(\beta_0=\log_{10}(\beta_0^*)\). Where relevant, to estimate the asymptotic rate of convergence, \(\beta_1\), we estimate the of the line \(\log_{10}(error) = \beta_0 + \beta_1 \log_{10}(p)\) using ordinary least squares and the last eight data points on each plot. We plot the estimated lines of best fit and quote their slope to the right of the plots.

% The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance. This allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics of the problem. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which allows us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates the ability of the schemes to approximate the limiting distribution of the model, Section~\ref{sec: transient approx} investigates the ability of the schemes to approximate the transient distribution of the model for two initial conditions, and Section~\ref{sec: return approx} investigates the ability of the schemes to approximate the first hitting time of the fluid level on the boundary of the interval \([0,1]\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num} and approximate the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{Y(t)\}\) first returns to \(0\).

@@

% 3. limiting DISTRIBUTIONS
	% i. SIMPLER
	% ii. MORE COMPLEX

@@

\section{Function approximation/reconstruction}\label{sec: recon num}
We start our investigation by looking at how well the schemes perform at approximating an initial condition. By approximating the initial condition only, we aim to investigate the performance of the approximation schemes without any dynamics.

For the discontinuous Galerkin method, we project the initial condition on to the set of polynomial basis functions which define the scheme, for the spatially-coherent uniformisation scheme, we look at the sub-cell averages of the initial condition, and for the QBD-RAP scheme, we compute the initial vector for the approximation, then reconstruct the solutions as described in Sections~\ref{sec: initial conditions} and~\ref{sec: closing}. For the purposes of approximation and reconstruction for the QBD-RAP scheme we must orientate the initial condition in a positive or negative direction; here we suppose that the initial conditions belong to a positive phase. First we investigate the three closing vectors from Section~\ref{sec: closing} which we can use for the reconstruction in the QBD-RAP scheme. From the investigation, we decide to use the closing vector (\ref{eqn:5462}) throughout the rest of the chapter.

In this section we consider approximating initial conditions over a single cell of width \(\Delta = 1\). To numerically evaluate integrals arising in the approximation step (inner products and cell averages) we use a trapezoidal rule with 10,001 function evaluations. Similarly, we use 10,001 function evaluations to approximate \(L^p\) errors, and also as a finite set of points over which to compute the Kolmogorov-Smirnov (KS) statistics.

@@

\subsection{QBD-RAP closing operators}\label{sec: closing vecs numerics}
The three closing operators we investigate are the following. The unnormalised closing vector, 
\begin{align}
	\bs a {\bs v}(x) &= \bs a e^{\bs Sx}\bs s.\tag{\ref{eqn: alal}}
	%\\\bs a {\bs v}_-(x-y_\ell) &= \bs a e^{\bs S(x-y_{\ell})}\bs s,\label{eqn:ndaenf}
\end{align} 
A normalised version of the closing operator in Equation~(\ref{eqn:546}), with the normalised version given by 
\begin{align}
	\bs a {\bs v}(x) &= \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e}. \tag{\ref{eqn: nonlin}}
\end{align} 
We refer to the closing operator in (\ref{eqn: nonlin}) as the \emph{non-linear normalised} closing operator.\footnote{Technically we have not yet proved that the \emph{non-linear} normalised closing operator leads to a convergent approximation scheme.} The third closing vector we investigate is given by 
\begin{align}
	\bs a {\bs v}(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)}\right)\left[I-e^{S2\Delta}\right]^{-1} \bs s,\tag{\ref{eqn:5462}}
\end{align}
which we refer to as the \emph{normalised} closing operator.

\begin{example}\label{ex: another example}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) of Example~\ref{ex: another example} for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves of the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}%To estimate the asymptotic rate of convergence, we fit a linear model of \(\log_{10}(error) = \beta_0 + \beta_1\log_{10}(Dim)\) to the the last 6 data points on the curves above, which corresponds to a rate of convergence \(error \propto N^{-\beta_1}\). The estimate of \(\beta_1\) was approximately -1.0 in all cases.}
		\label{fig: fun 2 ks error qbdrap closing vecs}
	\end{figure} 
	%\exampleFloatBarrier
	First consider approximating the initial condition with density function \(2\times 1(x<0.5)\), where \(1(\cdot)\) is the indicator function. In the left-hand side of Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} we plot the Kolmogorov-Smirnov error between the true CDF and the CDF constructed via the QBD-RAP approximation with the three different closing vectors. Interestingly, in this case, the unnormalised closing operator outperforms the other two at low dimensions. In this case, it just so happens that the error due to truncation of the unnormalised reconstruction constructively interferes with other errors to cause the error to be lower at low dimensions for the unnormalised scheme.

	Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also shows that the non-linear normalised and normalised reconstructions perform very similarly -- this is the case throughout much of this section. The difference between the non-linear normalised and normalised reconstructions is how they treat mass in the tail of a matrix exponential distribution (from \(2\Delta\) onward). Intuitively, there should be very little mass in the tail of the distribution as we are using concentrated matrix exponential distributions in the reconstruction.
